---
title: "Survival"
output: html_document
date: "2023-02-21"
---

## Color set

```{r}

library(devtools)
library(msSurv)
library(rmarkdown)
library(CoxHD)
library(mg14)
library(DT)
library(htmlwidgets)
library(tinytex)
library(dplyr)
library(mltools) #
library(data.table)


```

``` {r setup2, message=FALSE,warning=FALSE}
set1 <- brewer.pal(9, "Set1")
```


## 2.3.1 Number of oncogenic mutations

``` {r number_mut, message=FALSE,warning=FALSE}

#Define some indexes relating to subsets of variables used by the random effects model.

dataFrameOsTD <- dataFrame

mainGroups <- grep("[A-Z][a-z]+[A-Z]",levels(groups), invert=TRUE, value=TRUE)
mainGroups

mainIdx <- groups %in% mainGroups
osIdx <- !grepl("TPL", colnames(dataFrame)) ## Exclude TPL from OS analyses..
whichRFXOs <- which((colSums(dataFrame)>=MinOccurances | mainIdx) & osIdx) # ie, > 0.5%
mainIdxOs <- mainIdx & osIdx
osTDIdx <- !grepl("TPL_efs", colnames(dataFrame))
whichRFXOsTD <- which((colSums(dataFrame)>=MinOccurances | mainIdx) & osTDIdx) # ie, > 0.5%
mainIdxOsTD <- mainIdx & osTDIdx
whichRFXOsGG <- which((colSums(dataFrame)>=MinOccurances | mainIdxOs) & osIdx & 
                        groups %in% c(mainGroups,"GeneGene", "GeneCyto")) # ie, > 0.5%

#Compute the number of oncogenics, excluding complex karyotype.
NONC <- rowSums(cbind(dataList$Cytogenetics[names(dataList$Cytogenetics)!="complex"], 
                      dataList$Genetics), na.rm=TRUE)

NONCs <- factor(ceiling(pmin(NONC,6)/2), labels=c("0-1","2-3","4-6","7+"))
c <- set1[c(3,2,4,1,5)]
f <- survfit(osYr ~ NONCs)

```


## Survival plot
``` {r survival, message=FALSE,warning=FALSE}

s <- summary(f)
plot(f, col=c, xlim=c(0,10),xlab="Years", ylab="Survival",mark="|", cex=.5)
legend('topright', bty='n', col=c, legend=paste0(levels(NONCs)," (n=",table(NONCs),")"), lty=1)
```




## Linearity of continuous variables


``` {r spline, message=FALSE,warning=FALSE}

set.seed(42)
trainIdx <- sample(c(TRUE,FALSE), nrow(dataFrame), replace=TRUE, prob=c(0.66,0.34))
trainIdxOsTD <- trainIdx
par(mfrow=c(3,3))
clinicalSpline <- as.data.frame(sapply(dataFrame[groups %in% c("Clinical","Demographics")], function(x){
  if(all(x[1:5] %in% 0:10)) return(x)
  y <- log(x+min(x)+1e-3)
  fit <- coxph(os ~ pspline(y, df=3), subset=trainIdx)
  predict(fit, newdata=data.frame(y=y))
}))
for(n in names(clinicalSpline)) if(!all(dataFrame[1:5,n] %in% 0:10))
  plot(dataFrame[,n], clinicalSpline[,n], log='x', xlab=paste(n, '[observed]'), ylab = paste(n, '[spline]'))

```


## Calculate concordance

```{r concordance, message=FALSE, warning=FALSE}

summary(coxph(os ~ ., data=clinicalSpline, subset=!trainIdx))$concordance

summary(coxph(os ~ ., data=dataFrame[groups %in% c("Clinical","Demographics")]), subset=!trainIdx)$concordance

survConcordance(os[!trainIdx] ~ predict(coxph(os ~ ., data=clinicalSpline, subset=trainIdx),
                                        newdata = clinicalSpline[!trainIdx,]))
survConcordance(os[!trainIdx] ~ predict(coxph(os ~ ., data=dataFrame[groups %in% c("Clinical","Demographics")], 
                                              subset=trainIdx), newdata = dataFrame[!trainIdx,]))
```



# 2.3.2 Random effects models

```{r fitting, message=FALSE, warning=FALSE}

#Here we fit the random effects model using our implementation in the CoxHD package. First for main effects only
coxRFXFitOsTDMain <- CoxRFX(dataFrameOsTD[,mainIdxOsTD], osTD, groups[mainIdxOsTD], verbose = TRUE, max.iter = 100)
#Now including gene:gene interaction terms (min. recurrence = 8)
whichRFXOsTDGG <- which((colSums(dataFrame)>=MinOccurances | mainIdxOsTD) 
                        & osTDIdx & groups %in% c(mainGroups,"GeneGene")) 
coxRFXFitOsTDGGc <- CoxRFX(dataFrameOsTD[,whichRFXOsTDGG], osTD, groups[whichRFXOsTDGG], 
                           which.mu=mainGroups, verbose = TRUE, max.iter = 100) 
## allow only the main groups to have mean different from zero.. 


#Compute Harrel's concordance index

################## make sure it is finished running before continuing#################


```


## Stratified Cox Mode

```{r}
library(caret)

set.seed(123)

trainIndex <- createDataPartition(dataFrameOsTD$Protocol_Year, p = 0.6, list = FALSE, times = 1)

trainData <- dataFrameOsTD[trainIndex, ]
testData <- dataFrameOsTD[-trainIndex, ]

trainOsTD <- osTD[trainIndex]
testOsTD <- osTD[-trainIndex]


coxRFXFitOsTDMain <- CoxRFX(trainData[, mainIdxOsTD], trainOsTD, groups[mainIdxOsTD], verbose = TRUE, max.iter = 100)

whichRFXOsTDGG <- which((colSums(trainData) >= MinOccurances | mainIdxOsTD) 
                        & osTDIdx & groups %in% c(mainGroups, "GeneGene")) 

coxRFXFitOsTDGGc <- CoxRFX(trainData[, whichRFXOsTDGG], trainOsTD, groups[whichRFXOsTDGG], 
                           which.mu = mainGroups, verbose = TRUE, max.iter = 100)
```

more concordance
```{r concord2,message=FALSE, warning=FALSE}

survConcordance(trainOsTD~coxRFXFitOsTDGGc$linear.predictors)

```

```{r concord3,message=FALSE, warning=FALSE}
survConcordance(trainOsTD~coxRFXFitOsTDMain$linear.predictors)

```



# Variance components

Here we compute the variance components: 

```{r piechart_all, message=FALSE, warning=FALSE}

colGroups <- c("#E41A1C" , "#33A02C", "#FF7F00", "#FB9A99", "#A65628", "#6A3D9A", "#999999", "#A6CEE3", "#1F78B4", '#DAA520')

names(colGroups) <- levels(groups)[order(toupper(levels(groups)))]

Fig1<-PlotVarianceComponents(coxRFXFitOsTDGGc, col=colGroups,absolute = FALSE)
title("Risk contributions OS (time-dep) - 60% of the Data")

pdf(Fig1)
dev.print(pdf, file="Variance.pdf")

```


#2.3.3 Confidence intervals and significance tests
#Estimate confidence intervals by parametric bootstrap and compare with Wald Test. 
#Note that the usual sample with replacement yields inconsistencies for the interaction terms 
#due to the overdispersed correlations. The theoretical description of the survival time simulation
#is given in section survival.
###########################################***************************************************************
#TAKES A WHILE TO RUN
###########################################***************************************************************


```{r ci_etc, message=FALSE, warning=FALSE}


set.seed(42)
risk <- as.matrix(dataFrame[whichRFXOsTDGG]) %*% coxRFXFitOsTDGGc$coefficients
risk <- risk - mean(risk)


#########this takes time###################
parBoot <- mclapply(1:100, function(i) {
  s <- SimSurvNonp(risk, os)
  c <- try(CoxRFX(dataFrame[whichRFXOsTDGG], s, groups=groups[whichRFXOsTDGG], sigma0=0.1, nu=0))
  if(class(c)=="try-error")
    return(s)
  c$Z <- NULL # set X to zero to save mem
  return(c)
} )

```

```{r}

save(parBoot, file = "parBootNuis.Rda")
#load(file = "parBootNuis.Rda")


```


#Distributions of mean, sigma and df


```{r boxplot_sigma2, message=FALSE, warning=FALSE}
#Distributions of mean, sigma and df
boxplot(t(sapply(parBoot, `[[`, "sigma2")), border=colGroups[names(parBoot[[1]]$sigma2)], lty=1, pch=16, staplewex=0, ylab="sigma2", las=2, log="y", ylim=c(1e-3,8))
abline(h=0, lty=3)
points(coxRFXFitOsTDGGc$sigma2,  pch=19)
```



```{r boxplot_mu, message=FALSE, warning=FALSE}


boxplot(t(sapply(parBoot, `[[`, "mu")), border=colGroups[names(parBoot[[1]]$mu)], lty=1, pch=16, staplewex=0, ylab="mu", las=2)
abline(h=0, lty=3)
points(coxRFXFitOsTDGGc$mu,  pch=19)
```



```{r boxplot_df, message=FALSE, warning=FALSE}


boxplot(t(sapply(parBoot, `[[`, "df")), border=colGroups[names(parBoot[[1]]$mu)], lty=1, pch=16, staplewex=0, ylab="df", las=2)
abline(h=0, lty=3)
points(coxRFXFitOsTDGGc$df,  pch=19)
```


```{r bootstrap, message=FALSE, warning=FALSE}


#Coefficients
v <- apply(sapply(parBoot, `[[`, "coefficients"), 1, var, na.rm=TRUE)
w <- diag(coxRFXFitOsTDGGc$var) ## H^{-1}
w2 <- diag(coxRFXFitOsTDGGc$var2) ## H^{-1} I H^{-1}
c <- coef(coxRFXFitOsTDGGc)


plot(c^2/v, c^2/w, log="xy", xlab="Chi2 (bootstrap)", ylab="Chi2 (analyt.)", cex=.66)
par(xpd=NA)
points(c^2/v, c^2/w2, pch=16, cex=.7)
arrows(c^2/v, c^2/w, c^2/v,c^2/w2, length=0.05)
abline(0,1)
abline(h=qchisq(c(0.95,0.99,0.999), 1, lower.tail=TRUE), lty=c(1,2,3))
```



```{r wald, message=FALSE, warning=FALSE}


pBoot <- pchisq(c^2/v,1, lower.tail=FALSE)
pVar2 <- pchisq(c^2/w2,1, lower.tail=FALSE)
pVar <- pchisq(c^2/w,1, lower.tail=FALSE)


MainData <- dataRaw[whichRFXOsTDGG]



waldOut <- data.frame(group = groups[whichRFXOsTDGG], 
                      `beta (log-hazard)`= c, 
                      `hazard exp(beta)` = exp(c),
                      n = ifelse(groups[whichRFXOsTDGG] %in% c("CNA","Fusions","Genetics","GeneGene","Epitype", "Nuisance"), 
                                 colSums(dataRaw[sub("_10*$","",names(whichRFXOsTDGG))], na.rm=TRUE), NA),
                      
                      n2 = ifelse(groups[whichRFXOsTDGG] %in% c("Demographics","Clinical"), 
                                 colSums(MainData != "", na.rm=TRUE), NA),
                      

                  
                      
                      sd = sqrt(w2), 
                      `sd (bootstrap)` = sqrt(v), 
                      `sd (var)`= sqrt(w),
                      `P-value`=pVar2,
                      `Q (Benjamini-Yekutieli)` = p.adjust(pVar2, "BY"),
                      `Q (Benjamini-Hochberg)` = p.adjust(pVar2, "BH"),
                      check.names=FALSE
)


is.numeric(waldOut$n)
is.numeric(waldOut$n2)



waldOut$n2[is.na(waldOut$n2)] <- 0
waldOut$n[is.na(waldOut$n)] <- 0

waldOut$n <- waldOut$n + waldOut$n2

waldOut <- waldOut[,-5]



datatable(as.data.frame(lapply(waldOut, function(x) if(class(x)=="numeric") round(x,4) else x), 
                        check.names=FALSE, row.names=row.names(waldOut)))

waldOutDataTable <- as.data.frame(lapply(waldOut, function(x) if(class(x)=="numeric") round(x,4) else x), 
                                  check.names=FALSE, row.names=row.names(waldOut))

write.csv(waldOutDataTable, file= "Alliance with epi componant hazards min occurence 25 with Nuisance 2022.csv")




```




```{r volcano, message=FALSE, warning=FALSE}

#Volcano plot[]
par(mar=c(3,3,1,1)+.1,  bty="n", mgp=c(2,.5,0))
i <- coxRFXFitOsTDGGc$groups %in% c("Genetics", "CNA","Fusions","GeneGene","Epitype", "Nuisance")#apply(coxRFXFitOsTDGGc$Z,2,min) == 0 & apply(coxRFXFitOsTDGGc$Z,2,max) == 1
p <- pVar2 ## pvalues coxRFX

plot(c, 1/p, log='y', col=paste(colGroups[as.character(coxRFXFitOsTDGGc$groups)],"BB", sep=""), pch=ifelse(i,16,16), ylab="P-value",xlab="log hazard", cex=ifelse(i, sqrt(colMeans(coxRFXFitOsTDGGc$Z)*50),1), xlim=range(c*1.2))
abline(h=qchisq(c(0.95,0.99,0.999), 1, lower.tail=TRUE), lty=c(1,2,3))
w <- which(p.adjust(p,"BY") < 0.1)
points(c[w], 1/p[w],  pch=1, cex=ifelse(i[w], sqrt(colMeans(coxRFXFitOsTDGGc$Z[,w])*50),1))
w <- which(p.adjust(p,"bonf") < 0.05)
par(xpd=NA)
text(c[w], 1/p[w], names(c[w]), pos=3)
u <- par("usr")
f <- c(0.01,0.05,0.1,0.2,0.5)
s <- sqrt(f*50)
legend("topright",legend=f, pch=16, pt.cex=s, bty='n', col=paste("#88888888"))
par(xpd=FALSE)
abline(h=1/0.05, lty=2)
abline(h=1/max(p[which(p.adjust(p,"BY") < 0.1)]), lty=3)

```


```{r pvalue, message=FALSE, warning=FALSE}
#P-values and random model

set.seed(42)
Z <- apply(coxRFXFitOsTDGGc$Z, 2,sample)[1:nrow(dataFrame),] ## random covariates
coxRFXFitOsRain <- CoxRFX(Z, os, groups=coxRFXFitOsTDGGc$groups, nu=1) ## model
w2 <- diag(coxRFXFitOsRain$var2) 
c <- coef(coxRFXFitOsRain)
p2 <- pVar2

plot(seq(0,1,l=length(p2)+1)[-1],sort(p2), xlab="P-value (expected)", ylab="P-value (observed)", pch=16, col="grey")
abline(0,1)
points(seq(0,1,l=length(p)+1)[-1],sort(p), pch=16)
legend("topleft",bty="n", c("observed","randomised"), pch=16, col=c("black","grey"))

```

#Distribution of the variance components


```{r boxplot_var, message=FALSE, warning=FALSE}

v <- t(sapply(parBoot, function(x) {t <- try(VarianceComponents(x, newZ=dataFrame[whichRFXOsTDGG])); if(class(t)=="try-error") rep(NA, nlevels(x$groups)+1) else t}))
boxplot(v, border=colGroups[colnames(v)], lty=1, pch=16, staplewex=0, ylab="variance comp.", las=2)
abline(h=0, lty=3)
points(VarianceComponents(coxRFXFitOsTDGGc),  pch=19)

```


```{r loghazard, message=FALSE, warning=FALSE}
#Plot of log hazard v outcome
par(mar=c(3,3,3,1), mgp=c(2,.5,0))
t <- os
s <- survfit(os~1)
q <- quantile(t[,1], seq(0,1,.1))# q <- splinefun( s$surv, s$time,"monoH.FC")(seq(1,min(s$surv),l=10))
c <- cut(t[,1], q, na.rm=TRUE)
h <- coxRFXFitOsTDGGc$linear.predictors
o <- order(h)

plot(h[o], col= (brewer.pal(10,'RdBu'))[c[o]], type='h', xaxt="n", xlab='Patient', las=2, ylab="log hazard")
u <- par("usr")
q <- pmin(q,365*12)
image(x=q/max(q)*500, y=c(u[4]-(u[4]-u[3])/20, u[4]), matrix(1:10), col= (brewer.pal(10,'RdBu')), add=TRUE)
#axis(side=3, at=seq(1,500,l=11), labels=seq(0,1,.1))
axis(side=3, at=pretty(q/365)/max(q)*365*500, labels=pretty(q/365))
lines(ksmooth(seq_along(o),t[o,2]==0, bandwidth=50))

```
