---
title: "Models"
output: html_document
date: "2026-02-21"
---


###############################################
#Multistage predictor model
###############################################

Fits the multi-state disease trajectory model and separate CoxRFX models for each clinical endpoint (remission, early death, relapse, non-relapse death, post-relapse death). Includes variance component comparisons and per-endpoint volcano plots and Wald tables.

**Requires:** All objects from `01_data_preparation.Rmd` and `02_OS_model_inference.Rmd`.

```{r}
library(devtools)
library(msSurv)
library(rmarkdown)
library(CoxHD)
library(mg14)
library(DT)
library(htmlwidgets)
library(tinytex)
library(dplyr)
library(mltools) #
library(data.table)


```


```{r multistage, message=FALSE, warning=FALSE}

library(msSurv)


#outcome data
os_ms <- read.csv("endpoints_final.csv", header = TRUE, check.names = FALSE, stringsAsFactors = FALSE)
rownames(os_ms) <- os_ms$Sample
os_ms <- os_ms[,!(names(os_ms) %in% drop)]
os_new <- os_ms
for (i in 2:ncol(os_new)){os_new[,i] <- as.numeric(as.character(os_new[,i]))}


#try with as.numeric(os_new$Erdate[i]) = 0 - time for first visit - they changed from dates to time periods
#so recoding is needed and date isn't the correct thing to use.

d <- sapply(1:nrow(os_new), function(i){
  i <<- i
  t <- c(as.numeric(os_new[i,c("time_to_cr","time_to_relapse","time_to_fu")]) - 0)
  o <- order(t, na.last=NA)
  stages <- c(1:3,0)
  r <- stages[c(1, o+1)]
  if(os_new$sstat[i])
    r[length(r)] <- r[length(r)-1] +3
  tt <- c(0,t[o])
  if(length(o)==0)
    return(c(rep(NA,7),i))
  s <- cbind(id=i, stop=tt[-1], start.stage=r[-length(r)], end.stage=r[-1])[diff(tt)!=0,]
  #s <- cbind(time1 = tt[-length(tt)], time2=tt[-1], death=c(rep(0, length(o)-1), osDATA$Status[i]), outer(0:(length(o)-1), r[-3], `>=`)+0, i=i)[diff(tt)!=0,]
  return(s)
})

# had problems something with the erdate variable - made it zero which is what they have and now it works


d <- as.data.frame(do.call("rbind",d))
nodes <- as.character(1:6)
edges <- list(`1`=list(edges=c("2","4")), `2`=list(edges=c("3","5")), `3`=list(edges="6"), `4`=list(edges=NULL),
              `5`=list(edges=NULL),`6`=list(edges=NULL))
struct <- new("graphNEL", nodes = nodes, edgeL = edges, edgemode = "directed")
msurv <- msSurv(d, struct, bs = FALSE)
```



```{r diff_states, message=FALSE, warning=FALSE}
y <- t(apply(cbind(1,-msurv@ps[,c(4:6, 3:1)]),1,cumsum))
par(mar=c(3,3,1,1), bty="n", mgp=c(2,.5,0), las=1)
plot(msurv@et/365.25, y[,1], ylim=c(0,1), type="s",lty=0, xlab="Time after diagnosis", ylab="Fraction of patients", xlim=c(0,10), xaxs="i", yaxs="i")
steps <- function(x, type="s") rep(x, each=2)[if(type=="s") -1 else -2*length(x)]
x <- steps(msurv@et/365.25, type="S")
for(i in 1:6)
  polygon(c(x, rev(x)), c(steps(y[,i]), rev(steps(y[,i+1])) ), col=c(brewer.pal(5,"Pastel1")[c(1:3,5,4)],"#DDDDDD")[i], border=NA)
abline(h=seq(0,1,.2), col='white', lty=3)
abline(v=seq(0,10,1), col='white', lty=3)
lines(x, steps(y[,4]), lwd=2)
w <- which.min(abs(msurv@et/365.25-10))
text(x=par("usr")[2], y= y[w,-7]+diff(y[w,])/2, labels=c("early death","death in CR","death after relapse","alive with relapse","alive in remission","induction/LOF"), pos=2)

```

#setup for the multistage stuff so that we can get the individual end states plots


```{r diff_states_prep, message=FALSE, warning=FALSE}

library(CoxHD)
##TRANSPLANT STUFF

alloTimeCR1 <- os_new$time_to_cr + .5 # +.5 to make > 0
#alloTimeCR1[!alloIdx | (os_new$TPL_date < os_new$time_to_relapse & !os_new$TPL_Phase %in% c("CR1","RD"))] <- NA
whichRFXRel <- whichRFXOsTDGG[grep("TPL",names(whichRFXOsTDGG), invert=TRUE)] 
#mainIdx & !grepl("TPL", names(dataFrame)) & groups!="Nuisance"


t <- os_new$time_to_relapse
t[is.na(t)] <- as.Date(1e6, origin="2000-01-01")


relData <- MakeTimeDependent(dataFrame[whichRFXRel], timeEvent=alloTimeCR1,
                    timeStop=as.numeric(pmin(t, os_new$time_to_fu) - os_new$time_to_cr), 
                    status=!is.na(os_new$time_to_relapse)+0)

nrdData <- MakeTimeDependent(dataFrame[whichRFXRel], timeEvent=alloTimeCR1, 
                             timeStop=as.numeric(pmin(t, os_new$time_to_fu) - os_new$time_to_cr), 
                             status=is.na(os_new$time_to_relapse) & os_new$sstat)

alloTimeRel <- os_new$time_relapse + .5 # +.5 to make > 0

i <- !is.na(os_new$time_to_relapse)
prdData <- MakeTimeDependent(dataFrame[i,whichRFXRel], timeEvent=alloTimeRel[i],
                             timeStop=as.numeric(os_new$time_to_fu - os_new$time_to_relapse)[i], 
                             status=os_new$sstat[i])

w <- which(prdData$time1 == prdData$time2) ## 5 cases with LF=Rec
prdData$time2[w] <- prdData$time2[w] + .5


crGroups <- c(as.character(groups[whichRFXRel]))
names(crGroups) <- c(names(dataFrame)[whichRFXRel])


coxRFXNrdTD <- CoxRFX(nrdData[names(crGroups)], Surv(nrdData$time1, nrdData$time2, nrdData$status), 
                      groups=crGroups, nu = 1, which.mu = intersect(mainGroups, unique(crGroups)))


coxRFXPrdTD <- CoxRFX(prdData[names(crGroups)], Surv(prdData$time1, prdData$time2, prdData$status), 
                      groups=crGroups, nu=1, which.mu = intersect(mainGroups, unique(crGroups)))

coxRFXRelTD <- CoxRFX(relData[names(crGroups)], Surv(relData$time1, relData$time2, relData$status), 
                      groups=crGroups, nu = 1, which.mu = intersect(mainGroups, unique(crGroups)))

#added nu=1 to get coefficients that are not NA
#takes a few seconds to run above

osData2 <- MakeTimeDependent(dataFrame[whichRFXRel], timeEvent=alloTimeCR1, timeStop=as.numeric(os_new$time_to_fu - os_new$time_to_cr), status=os_new$sstat)

coxRFXOsCR <- CoxRFX(osData2[names(crGroups)], Surv(osData2$time1, osData2$time2, osData2$status), 
                     groups=crGroups, which.mu = intersect(mainGroups, unique(crGroups)))

table(CR=!is.na(os_new$time_to_cr), os[,2])
#c <- as.numeric(osDATA$time_to_cr - osDATA$Erdate)
c <- as.numeric(os_new$time_to_cr - 0)
c[is.na(c)] <- os_new$time_to_fu[is.na(c)]
cr <- Surv(c, factor(pmin(2 * (!is.na(os_new$time_to_cr))+os[,2],2), levels=0:2, 
                     labels=c("cens","ED","CR")), type="mstate")
coxRFXCrTD <- CoxRFX(osData2[1:1021, names(crGroups)], Surv(cr[,1], cr[,2]==2), groups=crGroups, nu=1,
                     which.mu = intersect(mainGroups, unique(crGroups)))
coxRFXNcdTD <- CoxRFX(osData2[1:1021, names(crGroups)], Surv(cr[,1], cr[,2]==1), groups=crGroups, which.mu = NULL, nu = 1)
#replaced which.mu=NULL with nu=1 to get coefficients that are not NA

```

#*Early Death pie chart

```{r pie_ed, message=FALSE, warning=FALSE}
o <- c(1,4,6,5,2,3,7,8)
PlotVarianceComponents(coxRFXNcdTD, col=colGroups, order=o)
title(main="Early deaths")

```

#*Remission pie chart

```{r pie_rem, message=FALSE, warning=FALSE}
PlotVarianceComponents(coxRFXCrTD, col=colGroups, order=o)
title(main="Remission")

```

#*Relapse pie chart

```{r Pie_relapse, message=FALSE, warning=FALSE}

PlotVarianceComponents(coxRFXRelTD, col=colGroups, order=o)
title(main="Relapse")


```


#*Non-relapse deaths pie chart

```{r pie_nonreldeath, message=FALSE, warning=FALSE}
PlotVarianceComponents(coxRFXNrdTD, col=colGroups, order=o)
title(main="Non-relapse deaths")

```


#*Post relapse death pie chart


```{r piepostreldeath, message=FALSE, warning=FALSE}
PlotVarianceComponents(coxRFXPrdTD, col=colGroups, order=o)
title(main="Post-relapse deaths")

```


```{r barplots1, message=FALSE, warning=FALSE}
par(mar=c(4,3,1,5))
allVarComp <- sapply(c("NcdTD","CrTD","NrdTD","RelTD","PrdTD"), function(x){
  m <- get(paste0("coxRFX",x))
  Z <- get(sub("\\[.+","",as.character(m$call["data"])))
  i <- if(x%in%c("CrTD","EsTD")) 1:1252 else Z$index
  VarianceComponents(m, newZ=Z[!rev(duplicated(rev(i))),colnames(m$Z)])})
colnames(allVarComp) <- c("Early deaths","Remission","Non-relapse d.","Relapse","Post-relapse d.")
w <- c("CNA","Fusions","Genetics","GeneGene","Clinical","Demographics","Epitype", "Nuisance")
z <- allVarComp[w,]/rep(colSums(allVarComp[-9,]), each=8)
b <- barplot(z, col=colGroups[w], ylab ="Variance [log hazard]", names.arg=rep("",ncol(z)))
rotatedLabel(x0=b, labels=colnames(z))
Z <- rbind(0,apply(z,2,cumsum))
n <- ncol(z)
segments(b[-n]+.5,t(Z[,-n]),b[-1]-.5 ,t(Z[,-1]))


```


```{r barplots2, message=FALSE, warning=FALSE}

par(mar=c(4,3,1,5))

colSums(allVarComp[-9,])


z <- allVarComp[w,]/rep(colSums(allVarComp[-9,]), each=8)


b <- barplot(z, col=colGroups[w], ylab="Relative importance", names.arg=rep("",ncol(z)))
rotatedLabel(x0=b, labels=colnames(z))
Z <- rbind(0,apply(z,2,cumsum))
n <- ncol(z)
segments(b[-n]+.5,t(Z[,-n]),b[-1]-.5 ,t(Z[,-1]))
mtext(side=4, at=Z[-1,n] - diff(Z[,n])/2, text=rownames(Z)[-1], las=2)

```


```{r barplots3, message=FALSE, warning=FALSE}

v <- c(1,3,5,4,2)
b <- barplot(z, col=colGroups[w], ylab="Relative importance", names.arg=rep("",ncol(z)))
rotatedLabel(x0=b, labels=colnames(z))
Z <- rbind(0,apply(z,2,cumsum))
n <- ncol(z)
segments(b[-n]+.5,t(Z[,-n]),b[-1]-.5 ,t(Z[,-1]))
mtext(side=4, at=Z[-1,n] - diff(Z[,n])/2, text=rownames(Z)[-1], las=2)



```




## Models concordance

```{r concordance2, messsage=FALSE, warning=FALSE}

coxRFXPrdTD$concordance

coxRFXNrdTD$concordance

coxRFXRelTD$concordance

coxRFXNcdTD$concordance

coxRFXOsCR$concordance

```


## Relapse data Volcano plot and table

```{r}

osRel <- Surv(relData$time1, relData$time2, relData$status)

relData2 <- relData[,names(coxRFXRelTD$coefficients)]

whichRFX <- which((colSums(relData2)>= 0))

set.seed(42)
risk <- as.matrix(relData2) %*% coxRFXRelTD$coefficients
risk <- risk - mean(risk)


SimSurvNonpS <- function(risk, surv, H0 = basehaz(coxph(surv ~ 1))) {
H0 = basehaz(coxph(os ~ 1))
FHazInv <- splinefun(c(0,H0$hazard), c(0,H0$time), method="monoH.FC", ties=min)
n = length(risk)
deathTimes = FHazInv(rexp(n, exp(risk))) #predict(hazardDist, rexp(n, exp(risk))) }

f <- os
f[,2] <- 1-f[,2]
F <- survfit(f~1)
FCensInv <- splinefun(F$surv, F$time, ties=min )
censTimes <- FCensInv(runif(n,0,1)) ## Simulate censoring times

survOut <-  Surv(time = pmax(0,pmin(deathTimes, censTimes)), event=(deathTimes < censTimes)+0)
return(survOut) }


#########this takes time###################
parBoot1 <- mclapply(1:100, function(i) {
  s <- SimSurvNonpS(risk, osRel)
  c <- try(CoxRFX(relData2[whichRFX], s, #groups=groups[whichRFX], sigma0=0.1, nu=0))
  if(class(c)=="try-error")
    return(s)
  c$Z <- NULL # set X to zero to save mem
  return(c)
 } )

save(parBoot1, file = "parBoot1Nuis.Rda")
#load(file = "parBoot1Nuis.Rda")

v <- apply(sapply(parBoot1, `[[`, "coefficients"), 1, var, na.rm=TRUE)
w <- diag(coxRFXRelTD$var) ## H^{-1}
w2 <- diag(coxRFXRelTD$var2) ## H^{-1} I H^{-1}
c <- coef(coxRFXRelTD)
plot(c^2/v, c^2/w, log="xy", xlab="Chi2 (bootstrap)", ylab="Chi2 (analyt.)", cex=.66)
par(xpd=NA)
points(c^2/v, c^2/w2, pch=16, cex=.7)
arrows(c^2/v, c^2/w, c^2/v,c^2/w2, length=0.05)
abline(0,1)
abline(h=qchisq(c(0.95,0.99,0.999), 1, lower.tail=TRUE), lty=c(1,2,3))

```


```{r rel2, message=FALSE, warning=FALSE}


#The plot indicates a good agreement of the variance estimate var2, see section 2.1.4.. Knowing the distribution of the variance allows us to compute a Wald test of the coefficients. #### Supplementary Table S1 Table with significance
pBoot <- pchisq(c^2/v,1, lower.tail=FALSE)
pVar2 <- pchisq(c^2/w2,1, lower.tail=FALSE)
pVar <- pchisq(c^2/w,1, lower.tail=FALSE)



Quack <- dataRaw[whichRFX]


waldOut1 <- data.frame(group = groups[whichRFX], 
                      `beta (log-hazard)`= c, 
                      `hazard exp(beta)` = exp(c),
                      n = ifelse(groups[whichRFX] %in% c("CNA","Fusions","Genetics","GeneGene","Epitype", "Nuisance"), 
                                 colSums(dataRaw[whichRFX], na.rm=TRUE), NA),
                      
                      n2 = ifelse(groups[whichRFX] %in% c("Demographics","Clinical"), 
                                 colSums(Quack != "", na.rm=TRUE), NA),
                      sd = sqrt(w2), 
                      `sd (bootstrap)` = sqrt(v), 
                      `sd (var)`= sqrt(w),
                      `P-value`=pVar2,
                      `Q (Benjamini-Yekutieli)` = p.adjust(pVar2, "BY"),
                      `Q (Benjamini-Hochberg)` = p.adjust(pVar2, "BH"),
                      check.names=FALSE
)



waldOut1$n2[is.na(waldOut1$n2)] <- 0
waldOut1$n[is.na(waldOut1$n)] <- 0

waldOut1$n <- waldOut1$n + waldOut1$n2

waldOut1 <- waldOut1[,-5]





datatable(as.data.frame(lapply(waldOut1, function(x) if(class(x)=="numeric") round(x,4) else x), check.names=FALSE
                        , row.names=row.names(waldOut1)))
waldOutDataTable1 <- as.data.frame(lapply(waldOut1, function(x) if(class(x)=="numeric") round(x,4) else x), check.names=FALSE, row.names=row.names(waldOut))


write.csv(waldOutDataTable1, file= "Relapse Data - WITH epi and nuisance - 2022.csv")
```

```{r}

w <- WaldTest(coxRFXRelTD)
w$Q.BH <- p.adjust(w$p.value, "BH")
w$Q.BY <- p.adjust(w$p.value, "BY")
datatable(w)

```

``` {r volcano relapse}
#Volcano plot

par(mar=c(3,3,1,1)+.1,  bty="n", mgp=c(2,.5,0))
par(mfrow=c(1,1))

i <- coxRFXRelTD$groups %in% c("Genetics", "CNA","Fusions","GeneGene","Epitype", "Nuisance")#apply(coxRFXFitOsTDGGc$Z,2,min) == 0 & apply(coxRFXFitOsTDGGc$Z,2,max) == 1
p <- pVar2 ## pvalues coxRFX
plot(c, 1/p, log='y', col=paste(colGroups[as.character(coxRFXRelTD$groups)],"BB", sep=""), pch=ifelse(i,16,16), ylab="P-value",xlab="log hazard", cex=ifelse(i, sqrt(colMeans(coxRFXFitOsTDGGc$Z)*50),1), xlim=range(c*1.2))
#abline(h=qchisq(c(0.95,0.99,0.999), 1, lower.tail=TRUE), lty=c(1,2,3))
w <- which(p.adjust(p,"fdr") < 0.1)
points(c[w], 1/p[w],  pch=1, cex=ifelse(i[w], sqrt(colMeans(coxRFXRelTD$Z[,w])*50),1))
w <- which(p.adjust(p,"fdr") < 0.05)
par(xpd=NA)
text(c[w], 1/p[w], names(c[w]), pos=3)
u <- par("usr")
f <- c(0.01,0.05,0.1,0.2,0.5)
s <- sqrt(f*50)
legend("topright",legend=f, pch=16, pt.cex=s, bty='n', col=paste("#88888888"))
par(xpd=FALSE)
abline(h=1/0.05, lty=2)
abline(h=1/max(p[which(p.adjust(p,"fdr") < 0.1)]), lty=3)

```


## Post relapse Data Volcano plot and Table 

```{r postrelapse, message=FALSE, warning=FALSE}

library(survival)
#this worked with the dataset wtih 1252 pts on 7/22/2021


osPrd <- survival::Surv(prdData$time1, prdData$time2, prdData$status)

prdData2 <- prdData[,names(coxRFXPrdTD$coefficients)]  #change columns to delete
mainGroups <- grep("[A-Z][a-z]+[A-Z]",levels(groups), invert=TRUE, value=TRUE)
mainGroups

mainIdx <- groups %in% mainGroups
whichRFX <- which((colSums(prdData2)>= 0)  )    #removed |MainIdx

set.seed(42)
risk <- as.matrix(prdData2) %*% coxRFXPrdTD$coefficients
risk <- risk - mean(risk)



SimSurvNonpS <- function(risk, surv, H0 = basehaz(coxph(surv ~ 1))) {
H0 = basehaz(coxph(os ~ 1))
FHazInv <- splinefun(c(0,H0$hazard), c(0,H0$time), method="monoH.FC", ties=min)
n = length(risk)
deathTimes = FHazInv(rexp(n, exp(risk))) #predict(hazardDist, rexp(n, exp(risk))) }

f <- os
f[,2] <- 1-f[,2]
F <- survfit(f~1)
FCensInv <- splinefun(F$surv, F$time, ties=min )
censTimes <- FCensInv(runif(n,0,1)) ## Simulate censoring times

survOut <-  Surv(time = pmax(0,pmin(deathTimes, censTimes)), event=(deathTimes < censTimes)+0)
return(survOut) }


parBoot2 <- mclapply(1:100, function(i) {
  s <- SimSurvNonpS(risk, osPrd)
  c <- try(CoxRFX(prdData2[whichRFX], s, #groups=groups[whichRFX], sigma0=0.1, nu=0))
  if(class(c)=="try-error")
    return(s)
  c$Z <- NULL # set X to zero to save mem
  return(c)
} )


save(parBoot2, file = "parBoot2Nuis.Rda")
#load(file = "parBoot2Nuis.Rda")

v <- apply(sapply(parBoot2, `[[`, "coefficients"), 1, var, na.rm=TRUE)
w <- diag(coxRFXPrdTD$var) ## H^{-1}
w2 <- diag(coxRFXPrdTD$var2) ## H^{-1} I H^{-1}
c <- coef(coxRFXPrdTD)
plot(c^2/v, c^2/w, log="xy", xlab="Chi2 (bootstrap)", ylab="Chi2 (analyt.)", cex=.66)
par(xpd=NA)
points(c^2/v, c^2/w2, pch=16, cex=.7)
arrows(c^2/v, c^2/w, c^2/v,c^2/w2, length=0.05)
abline(0,1)
abline(h=qchisq(c(0.95,0.99,0.999), 1, lower.tail=TRUE), lty=c(1,2,3))
```

``` {r wald postrelapse}

pBoot <- pchisq(c^2/v,1, lower.tail=FALSE)
pVar2 <- pchisq(c^2/w2,1, lower.tail=FALSE)
pVar <- pchisq(c^2/w,1, lower.tail=FALSE)

Quack <- dataRaw[whichRFX]


waldOut2 <- data.frame(group = groups[whichRFX], 
                      `beta (log-hazard)`= c, 
                      `hazard exp(beta)` = exp(c),
                      n = ifelse(groups[whichRFX] %in% c("CNA","Fusions","Genetics","GeneGene","Epitype", "Nuisance"), 
                                 colSums(dataRaw[whichRFX], na.rm=TRUE), NA),
                      
                      n2 = ifelse(groups[whichRFX] %in% c("Demographics","Clinical"), 
                                 colSums(MainData != "", na.rm=TRUE), NA),
                      sd = sqrt(w2), 
                      `sd (bootstrap)` = sqrt(v), 
                      `sd (var)`= sqrt(w),
                      `P-value`=pVar2,
                      `Q (Benjamini-Yekutieli)` = p.adjust(pVar2, "BY"),
                      `Q (Benjamini-Hochberg)` = p.adjust(pVar2, "BH"),
                      check.names=FALSE
)

waldOut2$n2[is.na(waldOut2$n2)] <- 0
waldOut2$n[is.na(waldOut2$n)] <- 0

waldOut2$n <- waldOut2$n + waldOut2$n2

waldOut2 <- waldOut2[,-5]




datatable(as.data.frame(lapply(waldOut2, function(x) if(class(x)=="numeric") round(x,4) else x), check.names=FALSE
                        , row.names=row.names(waldOut2)))
waldOutDataTable2 <- as.data.frame(lapply(waldOut2, function(x) if(class(x)=="numeric") round(x,4) else x), check.names=FALSE, row.names=row.names(waldOut2))


write.csv(waldOutDataTable2, file= "Postrelapse - WITH epi and Nuisance 2022.csv")
```


```{r}

w <- WaldTest(coxRFXPrdTD)
w$Q.BH <- p.adjust(w$p.value, "BH")
w$Q.BY <- p.adjust(w$p.value, "BY")
datatable(w)

```

``` {r volcano post relapse}
#Volcano plot
par(mar=c(3,3,1,1)+.1,  bty="n", mgp=c(2,.5,0))
par(mfrow=c(1,1))

i <- coxRFXPrdTD$groups %in% c("Genetics", "CNA","Fusions","GeneGene","Epitype", "Nuisance")#apply(coxRFXFitOsTDGGc$Z,2,min) == 0 & apply(coxRFXFitOsTDGGc$Z,2,max) == 1
p <- pVar2 ## pvalues coxRFX
plot(c, 1/p, log='y', col=paste(colGroups[as.character(coxRFXPrdTD$groups)],"BB", sep=""), pch=ifelse(i,16,16), ylab="P-value",xlab="log hazard", cex=ifelse(i, sqrt(colMeans(coxRFXPrdTD$Z)*50),1), xlim=range(c*1.2))
#abline(h=qchisq(c(0.95,0.99,0.999), 1, lower.tail=TRUE), lty=c(1,2,3))
w <- which(p.adjust(p,"fdr") < 0.1)
points(c[w], 1/p[w],  pch=1, cex=ifelse(i[w], sqrt(colMeans(coxRFXPrdTD$Z[,w])*50),1))
w <- which(p.adjust(p,"fdr") < 0.05)
par(xpd=NA)
text(c[w], 1/p[w], names(c[w]), pos=3)
u <- par("usr")
f <- c(0.01,0.05,0.1,0.2,0.5)
s <- sqrt(f*50)
legend("topright",legend=f, pch=16, pt.cex=s, bty='n', col=paste("#88888888"))
par(xpd=FALSE)
abline(h=1/0.05, lty=2)
abline(h=1/max(p[which(p.adjust(p,"fdr") < 0.1)]), lty=3)

```



## Complete Remission data volcano plot and table 

```{r CRdata , message=FALSE, warning=FALSE}

osCR <- survival::Surv(cr[,1], cr[,2]==2)
osData3 <- osData2[1:1021, names(crGroups)]

whichRFX <- which((colSums(osData3)>= 0)  )    #| mainIdx


set.seed(42)
risk <- as.matrix(osData3) %*% coxRFXCrTD$coefficients
risk <- risk - mean(risk)

SimSurvNonpS <- function(risk, surv, H0 = basehaz(coxph(surv ~ 1))) {
H0 = basehaz(coxph(os ~ 1))
FHazInv <- splinefun(c(0,H0$hazard), c(0,H0$time), method="monoH.FC", ties=min)
n = length(risk)
deathTimes = FHazInv(rexp(n, exp(risk))) #predict(hazardDist, rexp(n, exp(risk))) }

f <- os
f[,2] <- 1-f[,2]
F <- survfit(f~1)
FCensInv <- splinefun(F$surv, F$time, ties=min )
censTimes <- FCensInv(runif(n,0,1)) ## Simulate censoring times

survOut <-  Surv(time = pmax(0,pmin(deathTimes, censTimes)), event=(deathTimes < censTimes)+0)
return(survOut) }

parBoot3 <- mclapply(1:100, function(i) {
  s <- SimSurvNonpS(risk, osCR)
  c <- try(CoxRFX(osData3[whichRFX], s, groups=groups[whichRFX], #sigma0=0.1, nu=0))
  if(class(c)=="try-error")
    return(s)
  c$Z <- NULL # set X to zero to save mem
  return(c)
 } )

save(parBoot3, file = "parBoot3Nuis.Rda")
#load(file = "parBoot3Nuis.Rda")

v <- apply(sapply(parBoot3, `[[`, "coefficients"), 1, var, na.rm=TRUE)
w <- diag(coxRFXCrTD$var) ## H^{-1}
w2 <- diag(coxRFXCrTD$var2) ## H^{-1} I H^{-1}
c <- coef(coxRFXCrTD)
plot(c^2/v, c^2/w, log="xy", xlab="Chi2 (bootstrap)", ylab="Chi2 (analyt.)", cex=.66)
par(xpd=NA)
points(c^2/v, c^2/w2, pch=16, cex=.7)
arrows(c^2/v, c^2/w, c^2/v,c^2/w2, length=0.05)
abline(0,1)
abline(h=qchisq(c(0.95,0.99,0.999), 1, lower.tail=TRUE), lty=c(1,2,3))

```


``` {R wald CR}

pBoot <- pchisq(c^2/v,1, lower.tail=FALSE)
pVar2 <- pchisq(c^2/w2,1, lower.tail=FALSE)
pVar <- pchisq(c^2/w,1, lower.tail=FALSE)

Quack <- dataRaw[whichRFX]


waldOut3 <- data.frame(group = groups[whichRFX], 
                      `beta (log-hazard)`= c, 
                      `hazard exp(beta)` = exp(c),
                      n = ifelse(groups[whichRFX] %in% c("CNA","Fusions","Genetics","GeneGene","Epitype", "Nuisance"), 
                                 colSums(dataRaw[whichRFX], na.rm=TRUE), NA),
                      
                      n2 = ifelse(groups[whichRFX] %in% c("Demographics","Clinical"), 
                                 colSums(Quack != "", na.rm=TRUE), NA),
                      sd = sqrt(w2), 
                      `sd (bootstrap)` = sqrt(v), 
                      `sd (var)`= sqrt(w),
                      `P-value`=pVar2,
                      `Q (Benjamini-Yekutieli)` = p.adjust(pVar2, "BY"),
                      `Q (Benjamini-Hochberg)` = p.adjust(pVar2, "BH"),
                      check.names=FALSE
)


waldOut3$n2[is.na(waldOut3$n2)] <- 0
waldOut3$n[is.na(waldOut3$n)] <- 0

waldOut3$n <- waldOut3$n + waldOut3$n2

waldOut3 <- waldOut3[,-5]


datatable(as.data.frame(lapply(waldOut3, function(x) if(class(x)=="numeric") round(x,4) else x), check.names=FALSE
                        , row.names=row.names(waldOut3)))
waldOutDataTable3 <- as.data.frame(lapply(waldOut3, function(x) if(class(x)=="numeric") round(x,4) else x), check.names=FALSE, row.names=row.names(waldOut3))

write.csv(waldOutDataTable3, file= "CR data - WITH epi with Nuisance - 2022.csv")

```

```{r}

w <- WaldTest(coxRFXCrTD)
w$Q.BH <- p.adjust(w$p.value, "BH")
w$Q.BY <- p.adjust(w$p.value, "BY")
datatable(w)

```

``` {R CR volcano}


#Volcano plot
par(mar=c(3,3,1,1)+.1,  bty="n", mgp=c(2,.5,0))
par(mfrow=c(1,1))

i <- coxRFXCrTD$groups %in% c("Genetics", "CNA","Fusions","GeneGene","Epitype", "Nuisance")
p <- pVar2 ## pvalues coxRFX
plot(c, 1/p, log='y', col=paste(colGroups[as.character(coxRFXCrTD$groups)],"BB", sep=""), pch=ifelse(i,16,16), ylab="P-value",xlab="log hazard", cex=ifelse(i, sqrt(colMeans(coxRFXCrTD$Z)*50),1), xlim=range(c*1.2))
#abline(h=qchisq(c(0.95,0.99,0.999), 1, lower.tail=TRUE), lty=c(1,2,3))
w <- which(p.adjust(p,"fdr") < 0.1)
points(c[w], 1/p[w],  pch=1, cex=ifelse(i[w], sqrt(colMeans(coxRFXCrTD$Z[,w])*50),1))
w <- which(p.adjust(p,"fdr") < 0.05)
par(xpd=NA)
text(c[w], 1/p[w], names(c[w]), pos=3)
u <- par("usr")
f <- c(0.01,0.05,0.1,0.2,0.5)
s <- sqrt(f*50)
legend("topright",legend=f, pch=16, pt.cex=s, bty='n', col=paste("#88888888"))
par(xpd=FALSE)
abline(h=1/0.05, lty=2)
abline(h=1/max(p[which(p.adjust(p,"fdr") < 0.1)]), lty=3)

```


## Non-complete remission death 

```{r Ncddata, message=FALSE, warning=FALSE}

library(survival)


osNcd <- survival::Surv(cr[,1], cr[,2]==1)
osData3 <- osData2[1:1021, names(crGroups)]

whichRFX <- which((colSums(osData3)>= 0))



set.seed(42)
risk <- as.matrix(osData3) %*% coxRFXNcdTD$coefficients
risk <- risk - mean(risk)

SimSurvNonpS <- function(risk, surv, H0 = basehaz(coxph(surv ~ 1))) {
H0 = basehaz(coxph(os ~ 1))
FHazInv <- splinefun(c(0,H0$hazard), c(0,H0$time), method="monoH.FC", ties=min)
n = length(risk)
deathTimes = FHazInv(rexp(n, exp(risk))) #predict(hazardDist, rexp(n, exp(risk))) }

f <- os
f[,2] <- 1-f[,2]
F <- survfit(f~1)
FCensInv <- splinefun(F$surv, F$time, ties=min )
censTimes <- FCensInv(runif(n,0,1)) ## Simulate censoring times

survOut <-  Surv(time = pmax(0,pmin(deathTimes, censTimes)), event=(deathTimes < censTimes)+0)
return(survOut) }

parBoot4 <- mclapply(1:100, function(i) {
  s <- SimSurvNonpS(risk, osNcd)
  c <- try(CoxRFX(osData3[whichRFX], s, groups=groups[whichRFX], sigma0=0.1, nu=0))
  if(class(c)=="try-error")
    return(s)
  c$Z <- NULL # set X to zero to save mem
  return(c)
 } )

save(parBoot4, file = "parBoot4Nuis.Rda")
#load(file = "parBoot4Nuis.Rda")



v <- apply(sapply(parBoot4, `[[`, "coefficients"), 1, var, na.rm=TRUE)
w <- diag(coxRFXNcdTD$var) ## H^{-1}
w2 <- diag(coxRFXNcdTD$var2) ## H^{-1} I H^{-1}
c <- coef(coxRFXNcdTD)
plot(c^2/v, c^2/w, log="xy", xlab="Chi2 (bootstrap)", ylab="Chi2 (analyt.)", cex=.66)
par(xpd=NA)
points(c^2/v, c^2/w2, pch=16, cex=.7)
arrows(c^2/v, c^2/w, c^2/v,c^2/w2, length=0.05)
abline(0,1)
abline(h=qchisq(c(0.95,0.99,0.999), 1, lower.tail=TRUE), lty=c(1,2,3))

pBoot <- pchisq(c^2/v,1, lower.tail=FALSE)
pVar2 <- pchisq(c^2/w2,1, lower.tail=FALSE)
pVar <- pchisq(c^2/w,1, lower.tail=FALSE)
```

```{r Ncd2 wald, message=FALSE, warning=FALSE}

Quack <- dataRaw[whichRFX]


waldOut4 <- data.frame(group = groups[whichRFX], 
                      `beta (log-hazard)`= c, 
                      `hazard exp(beta)` = exp(c),
                      n = ifelse(groups[whichRFX] %in% c("CNA","Fusions","Genetics","GeneGene","Epitype", "Nuisance"), 
                                 colSums(dataRaw[whichRFX], na.rm=TRUE), NA),
                      
                      n2 = ifelse(groups[whichRFX] %in% c("Demographics","Clinical"), 
                                 colSums(Quack != "", na.rm=TRUE), NA),
                      sd = sqrt(w2), 
                      `sd (bootstrap)` = sqrt(v), 
                      `sd (var)`= sqrt(w),
                      `P-value`=pVar2,
                      `Q (Benjamini-Yekutieli)` = p.adjust(pVar2, "BY"),
                      `Q (Benjamini-Hochberg)` = p.adjust(pVar2, "BH"),
                      check.names=FALSE
)



waldOut4$n2[is.na(waldOut4$n2)] <- 0
waldOut4$n[is.na(waldOut4$n)] <- 0

waldOut4$n <- waldOut4$n + waldOut4$n2

waldOut4 <- waldOut4[,-5]






datatable(as.data.frame(lapply(waldOut4, function(x) if(class(x)=="numeric") round(x,4) else x), check.names=FALSE
                        , row.names=row.names(waldOut4)))
waldOutDataTable4 <- as.data.frame(lapply(waldOut4, function(x) if(class(x)=="numeric") round(x,4) else x), check.names=FALSE, row.names=row.names(waldOut4))

write.csv(waldOutDataTable4, file= "Ncd data - WITH epi and Nuisance - 2022.csv")
```


``` {R NCD volcano}
#Volcano plot
par(mar=c(3,3,1,1)+.1,  bty="n", mgp=c(2,.5,0))
par(mfrow=c(1,1))

i <- coxRFXNcdTD$groups %in% c("Genetics", "CNA","Fusions","GeneGene","Epitype", "Nuisance")
p <- pVar2 ## pvalues coxRFX
plot(c, 1/p, log='y', col=paste(colGroups[as.character(coxRFXNcdTD$groups)],"BB", sep=""), pch=ifelse(i,16,16), ylab="P-value",xlab="log hazard", cex=ifelse(i, sqrt(colMeans(coxRFXNcdTD$Z)*50),1), xlim=range(c*1.2))
#abline(h=qchisq(c(0.95,0.99,0.999), 1, lower.tail=TRUE), lty=c(1,2,3))
w <- which(p.adjust(p,"fdr") < 0.1)
points(c[w], 1/p[w],  pch=1, cex=ifelse(i[w], sqrt(colMeans(coxRFXNcdTD$Z[,w])*50),1))
w <- which(p.adjust(p,"fdr") < 0.05)
par(xpd=NA)
text(c[w], 1/p[w], names(c[w]), pos=3)
u <- par("usr")
f <- c(0.01,0.05,0.1,0.2,0.5)
s <- sqrt(f*50)
legend("topright",legend=f, pch=16, pt.cex=s, bty='n', col=paste("#88888888"))
par(xpd=FALSE)
abline(h=1/0.05, lty=2)
abline(h=1/max(p[which(p.adjust(p,"fdr") < 0.1)]), lty=3)

```



#Nonrelapse death


```{r Nrd, message=FALSE, warning=FALSE}

osNrd <- Surv(nrdData$time1, nrdData$time2, nrdData$status)

nrdData2 <- nrdData[,names(coxRFXNrdTD$coefficients)]

whichRFX <- which((colSums(nrdData2)>= 0) )

set.seed(42)
risk <- as.matrix(nrdData2) %*% coxRFXNrdTD$coefficients
risk <- risk - mean(risk)


SimSurvNonpS <- function(risk, surv, H0 = basehaz(coxph(surv ~ 1))) {
H0 = basehaz(coxph(os ~ 1))
FHazInv <- splinefun(c(0,H0$hazard), c(0,H0$time), method="monoH.FC", ties=min)
n = length(risk)
deathTimes = FHazInv(rexp(n, exp(risk))) #predict(hazardDist, rexp(n, exp(risk))) }

f <- os
f[,2] <- 1-f[,2]
F <- survfit(f~1)
FCensInv <- splinefun(F$surv, F$time, ties=min )
censTimes <- FCensInv(runif(n,0,1)) ## Simulate censoring times

survOut <-  Surv(time = pmax(0,pmin(deathTimes, censTimes)), event=(deathTimes < censTimes)+0)
return(survOut) }


parBoot5 <- mclapply(1:100, function(i) {
  s <- SimSurvNonpS(risk, osNrd)
  c <- try(CoxRFX(nrdData2[whichRFX], s, #groups=groups[whichRFX], sigma0=0.1, nu=0))
  if(class(c)=="try-error")
    return(s)
  c$Z <- NULL # set X to zero to save mem
  return(c)
} )

save(parBoot5, file = "parBoot5Nuis.Rda")
#load(file = "parBoot5Nuis.Rda")

```

``` {r NRD}

v <- apply(sapply(parBoot5, `[[`, "coefficients"), 1, var, na.rm=TRUE)
w <- diag(coxRFXNrdTD$var) ## H^{-1}
w2 <- diag(coxRFXNrdTD$var2) ## H^{-1} I H^{-1}
c <- coef(coxRFXNrdTD)
plot(c^2/v, c^2/w, log="xy", xlab="Chi2 (bootstrap)", ylab="Chi2 (analyt.)", cex=.66)
par(xpd=NA)
points(c^2/v, c^2/w2, pch=16, cex=.7)
arrows(c^2/v, c^2/w, c^2/v,c^2/w2, length=0.05)
abline(0,1)
abline(h=qchisq(c(0.95,0.99,0.999), 1, lower.tail=TRUE), lty=c(1,2,3))
```

``` {r NRD wald }
pBoot <- pchisq(c^2/v,1, lower.tail=FALSE)
pVar2 <- pchisq(c^2/w2,1, lower.tail=FALSE)
pVar <- pchisq(c^2/w,1, lower.tail=FALSE)

Quack <- dataRaw[whichRFX]

#Clinical and Demographic aren't in c bc the n is large but all the pvalue info is in the table.
waldOut5 <- data.frame(group = groups[whichRFX], 
                      `beta (log-hazard)`= c, 
                      `hazard exp(beta)` = exp(c),
                      n = ifelse(groups[whichRFX] %in% c("CNA","Fusions","Genetics","GeneGene","Epitype", "Nuisance"), 
                                 colSums(dataRaw[whichRFXRel], na.rm=TRUE), NA),
                      
                      n2 = ifelse(groups[whichRFX] %in% c("Demographics","Clinical"), 
                                 colSums(Quack != "", na.rm=TRUE), NA),
                      sd = sqrt(w2), 
                      `sd (bootstrap)` = sqrt(v), 
                      `sd (var)`= sqrt(w),
                      `P-value`=pVar2,
                      `Q (Benjamini-Yekutieli)` = p.adjust(pVar2, "BY"),
                      `Q (Benjamini-Hochberg)` = p.adjust(pVar2, "BH"),
                      check.names=FALSE
)


waldOut5$n2[is.na(waldOut5$n2)] <- 0
waldOut5$n[is.na(waldOut5$n)] <- 0

waldOut5$n <- waldOut5$n + waldOut5$n2

waldOut5 <- waldOut5[,-5]




datatable(as.data.frame(lapply(waldOut5, function(x) if(class(x)=="numeric") round(x,4) else x), check.names=FALSE
                        , row.names=row.names(waldOut5)))
waldOutDataTable5 <- as.data.frame(lapply(waldOut5, function(x) if(class(x)=="numeric") round(x,4) else x), check.names=FALSE, row.names=row.names(waldOut5))

write.csv(waldOutDataTable5, file= "Nrd Data - WITH epi and Nuisance -2022.csv")
```

``` {R NRD volcano}
#Volcano plot
par(mar=c(3,3,1,1)+.1,  bty="n", mgp=c(2,.5,0))
par(mfrow=c(1,1))

i <- coxRFXNrdTD$groups %in% c("Genetics", "CNA","Fusions","GeneGene","Epitype", "Nuisance")
p <- pVar2 ## pvalues coxRFX
plot(c, 1/p, log='y', col=paste(colGroups[as.character(coxRFXNrdTD$groups)],"BB", sep=""), pch=ifelse(i,16,16), ylab="P-value",xlab="log hazard", cex=ifelse(i, sqrt(colMeans(coxRFXNrdTD$Z)*50),1), xlim=range(c*1.2))
#abline(h=qchisq(c(0.95,0.99,0.999), 1, lower.tail=TRUE), lty=c(1,2,3))
w <- which(p.adjust(p,"fdr") < 0.1)
points(c[w], 1/p[w],  pch=1, cex=ifelse(i[w], sqrt(colMeans(coxRFXNrdTD$Z[,w])*50),1))
w <- which(p.adjust(p,"fdr") < 0.05)
par(xpd=NA)
text(c[w], 1/p[w], names(c[w]), pos=3)
u <- par("usr")
f <- c(0.01,0.05,0.1,0.2,0.5)
s <- sqrt(f*50)
legend("topright",legend=f, pch=16, pt.cex=s, bty='n', col=paste("#88888888"))
par(xpd=FALSE)
abline(h=1/0.05, lty=2)
abline(h=1/max(p[which(p.adjust(p,"fdr") < 0.1)]), lty=3)



```

